---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%"
)
```

# ICtoolkit

<!-- badges: start -->
[![R-CMD-check](https://github.com/petersonR/ICtoolkit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/petersonR/ICtoolkit/actions/workflows/R-CMD-check.yaml)
[![Codecov](https://codecov.io/gh/petersonR/ICtoolkit/branch/main/graph/badge.svg)](https://app.codecov.io/gh/petersonR/ICtoolkit)
<!-- badges: end -->

ICtoolkit provides a unified interface for computing five information criteria
across four model classes, plus a flexible stepwise selection function.

| Criterion | Function | Reference |
|-----------|----------|-----------|
| AIC | `compute_aic()` | Akaike (1974) |
| AICc | `compute_aicc()` | Hurvich & Tsai (1989) |
| BIC | `compute_bic()` | Schwarz (1978) |
| EBIC | `compute_ebic()` | Chen & Chen (2008) |
| RBIC | `compute_rbic()` | Peterson & Cavanaugh (2026+) |

**Supported model classes:** `lm`, `glm`, `glmnet`, `ncvreg`.
For `glmnet` and `ncvreg`, each criterion returns a vector of length
`length(lambda)` — one value per point on the regularization path.

## Installation

```{r, eval = FALSE}
# Development version from this repository:
devtools::install_github("petersonR/ICtoolkit")

# Or once on CRAN:
install.packages("ICtoolkit")
```

## Quick start

### IC for `lm` / `glm`

```{r lm-quick}
library(ICtoolkit)

fit <- lm(mpg ~ wt + hp + cyl, data = mtcars)
compute_aic(fit)
compute_bic(fit)
compute_ebic(fit, P = 10)   # P = total candidate predictors
```

Results carry informative attributes:

```{r attrs}
result <- compute_bic(fit)
attr(result, "fit_class")
attr(result, "k")
attr(result, "criterion")
```

### IC over the regularization path (`glmnet`)

```{r glmnet-quick, eval = requireNamespace("glmnet", quietly = TRUE)}
library(glmnet)

n <- 100
p <- 50

set.seed(1)

X <- matrix(rnorm(n * p), n, p,
            dimnames = list(NULL, paste0("x", 1:p)))
y <- X[, 1] * 2 - X[, 3] * 0.25 + rnorm(n)
fit_glmnet <- glmnet(X, y)

# plot_ic_path computes + plots IC paths in one call.
plot_ic_path(fit_glmnet, criteria = c("AIC", "BIC", "EBIC"),
             main = "IC paths for glmnet")
```

### RBIC for grouped predictors

RBIC applies group-specific EBIC penalties — useful when predictors have a
natural structure (e.g., main effects vs. interactions, or different data
modalities):

```{r rbic-quick}
P_index <- list(
  main         = c("wt", "hp", "disp"),
  categorical  = c("cyl", "gear", "carb", "am", "vs")
)
compute_rbic(fit, P_index = P_index)
```

### Stepwise selection with `ic_step()`

`ic_step()` generalises `MASS::stepAIC` to all five criteria:

```{r step-quick}
fit_null <- lm(mpg ~ 1, data = mtcars)
fit_full <- lm(mpg ~ ., data = mtcars)

selected <- ic_step(
  fit_null,
  scope     = list(lower = fit_null, upper = fit_full),
  direction = "both",
  criterion = "BIC",
  trace     = 0
)
formula(selected)

# The selection path is stored as an attribute
attr(selected, "step_path")
```

This package was developed by Ryan Peterson. Code review, parallelization, and testing were assisted by Claude (Anthropic).
