---
title: "Getting started with ICtoolkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ICtoolkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

```{r load}
library(ICtoolkit)
```

## Overview

ICtoolkit provides a unified interface for five information criteria across four
model classes:

| Criterion | Function | Description |
|-----------|----------|-------------|
| AIC | `compute_aic()` | Akaike (1974) |
| AICc | `compute_aicc()` | Small-sample correction |
| BIC | `compute_bic()` | Schwarz (1978) |
| EBIC | `compute_ebic()` | Extended BIC; Chen & Chen (2008) |
| RBIC | `compute_rbic()` | Grouped EBIC for structured predictors |

All functions support `lm`, `glm`, `glmnet`, and `ncvreg` objects.
For `glmnet` and `ncvreg`, the criterion is evaluated at every point on the
regularization path (one value per `lambda`), which can itself be used as a
selection algorithm without any stepwise fitting.

---

## 1. Standard criteria for `lm` and `glm`

```{r lm-example}
fit <- lm(mpg ~ wt + hp + cyl, data = mtcars)

compute_aic(fit)
compute_aicc(fit)
compute_bic(fit)
```

Each result is a scalar with informative attributes:

```{r attributes}
result <- compute_bic(fit)
attr(result, "fit_class")   # model class
attr(result, "k")           # number of estimated parameters
attr(result, "criterion")   # which IC
```

---

## 2. EBIC: Extended BIC

EBIC adds a sparsity-encouraging penalty to BIC:

$$\text{EBIC} = \text{BIC} + 2\,\gamma\,\log\binom{P}{k}$$

where $P$ is the total number of candidate predictors and $k$ is the number
selected.  The default $\gamma$ follows Chen & Chen (2008):

$$\gamma = \frac{\log(P/\sqrt{n})}{\log P} \cdot \mathbf{1}(P \geq \sqrt{n})$$

```{r ebic-lm}
# P = 10 candidate predictors (total, excluding intercept)
result_ebic <- compute_ebic(fit, P = 10)
result_ebic

attr(result_ebic, "gamma")   # the computed gamma
attr(result_ebic, "P")
```

You can supply your own `gammafn`:

```{r ebic-custom-gamma}
# Fixed gamma = 0.5
compute_ebic(fit, P = 10, gammafn = function(P, k, n) 0.5)
```

Setting `gammafn = function(P, k, n) 0` recovers standard BIC:

```{r ebic-zero-gamma}
all.equal(
  as.numeric(compute_ebic(fit, P = 10, gammafn = function(P, k, n) 0)),
  as.numeric(compute_bic(fit))
)
```

---

## 3. RBIC: Grouped EBIC for structured predictors

RBIC generalises EBIC to predictors with a natural group structure (e.g.,
main effects vs. interactions, or predictors from different data modalities):

$$\text{RBIC} = \text{BIC} + 2 \sum_{g=1}^{G} \gamma_g \log\binom{P_g}{k_g}$$

Groups are defined via a `P_index` list.  For `lm`/`glm` objects, each
element should be a character vector of **term labels** exactly as returned
by `attr(terms(fit), "term.labels")`.

```{r rbic-setup}
set.seed(1)
n  <- 100
df <- data.frame(
  y    = rnorm(n),
  x1   = rnorm(n), x2 = rnorm(n), x3 = rnorm(n),   # main effects
  x1sq = rnorm(n), x2sq = rnorm(n)                   # squared terms
)
fit2 <- lm(y ~ x1 + x2 + x3 + x1sq + x2sq, data = df)

P_index <- list(
  main    = c("x1", "x2", "x3"),
  squared = c("x1sq", "x2sq")
)

result_rbic <- compute_rbic(fit2, P_index = P_index)
result_rbic

attr(result_rbic, "gamma")    # per-group gamma
```

---

## 4. IC over the regularization path with `glmnet`

For `glmnet` objects, all five criteria return a **vector** of length
`length(lambda)`.  This makes IC-based selection straightforward: pick the
`lambda` that minimises the criterion.

```{r glmnet-example, eval = requireNamespace("glmnet", quietly = TRUE)}
library(glmnet)
set.seed(42)
n  <- 200; p <- 20
X  <- matrix(rnorm(n * p), n, p,
             dimnames = list(NULL, paste0("x", 1:p)))
y  <- X[, 1] * 2 - X[, 3] + X[, 7] * 0.5 + rnorm(n)

fit_glmnet <- glmnet(X, y, family = "gaussian")

bic_path  <- compute_bic(fit_glmnet)
ebic_path <- compute_ebic(fit_glmnet, P = p)
rbic_path <- compute_rbic(fit_glmnet,
                           P_index = list(first10  = paste0("x", 1:10),
                                          second10 = paste0("x", 11:20)))

# Lambda selected by each criterion
lambda_bic  <- attr(bic_path,  "lambda")[which.min(bic_path)]
lambda_ebic <- attr(ebic_path, "lambda")[which.min(ebic_path)]
lambda_rbic <- attr(rbic_path, "lambda")[which.min(rbic_path)]

cat("BIC-selected lambda: ",  round(lambda_bic,  4), "\n")
cat("EBIC-selected lambda:", round(lambda_ebic, 4), "\n")
cat("RBIC-selected lambda:", round(lambda_rbic, 4), "\n")

# Coefficients at BIC-optimal lambda
coef(fit_glmnet, s = lambda_bic)
```

> **Note on alpha < 1:** When a `glmnet` model is fit with `alpha < 1`
> (elastic-net or ridge), the package issues a warning because the count of
> nonzero coefficients may not be an appropriate measure of model complexity
> under L2 shrinkage.

---

## 5. Stepwise selection with `ic_step()`

`ic_step()` extends `MASS::stepAIC` to all five criteria for `lm`/`glm`
objects.

```{r ic-step-example}
fit_null <- lm(mpg ~ 1,  data = mtcars)
fit_full <- lm(mpg ~ .,  data = mtcars)

# Forward BIC selection
selected <- ic_step(
  fit_null,
  scope     = list(lower = fit_null, upper = fit_full),
  direction = "forward",
  criterion = "BIC",
  trace     = 1
)
formula(selected)
```

### Inspecting the selection path

`ic_step()` attaches a `step_path` data frame to the returned model:

```{r step-path}
attr(selected, "step_path")
```

### RBIC stepwise selection

For RBIC, supply `P_index` with term labels that match the scope's upper model:

```{r rbic-step, eval = FALSE}
P_index <- list(
  continuous   = c("wt", "hp", "disp", "drat", "qsec"),
  categorical  = c("cyl", "gear", "carb", "am", "vs")
)

ic_step(
  fit_null,
  scope     = list(lower = fit_null, upper = fit_full),
  direction = "forward",
  criterion = "RBIC",
  P_index   = P_index,
  trace     = 1
)
```

---

## 6. Comparing criteria on simulated data

```{r comparison}
set.seed(2026)
n <- 80; p_true <- 3; p_total <- 15
X_mat <- matrix(rnorm(n * p_total), n, p_total,
                dimnames = list(NULL, paste0("x", 1:p_total)))
beta  <- c(2, -1.5, 1, rep(0, p_total - p_true))
y_sim <- X_mat %*% beta + rnorm(n, sd = 2)
df_sim <- as.data.frame(cbind(y = y_sim, X_mat))

fit_null_sim <- lm(y ~ 1,  data = df_sim)
fit_full_sim <- lm(y ~ .,  data = df_sim)

criteria <- c("AIC", "AICc", "BIC", "EBIC", "RBIC")
results  <- lapply(criteria, function(crit) {
  extra_args <- list()
  if (crit == "EBIC") extra_args <- list(P = p_total)
  if (crit == "RBIC") extra_args <- list(
    P_index = list(first5 = paste0("x", 1:5),
                   rest   = paste0("x", 6:p_total))
  )
  do.call(ic_step, c(
    list(object    = fit_null_sim,
         scope     = list(lower = fit_null_sim, upper = fit_full_sim),
         direction = "forward",
         criterion = crit,
         trace     = 0),
    extra_args
  ))
})
names(results) <- criteria

# Number of terms selected by each criterion
sapply(results, function(m) length(attr(terms(m), "term.labels")))
```

---

## References

Akaike, H. (1974). A new look at the statistical model identification.
*IEEE Transactions on Automatic Control*, **19**(6), 716–723.

Chen, J., & Chen, Z. (2008). Extended Bayesian information criteria for model
selection with large model spaces. *Biometrika*, **95**(3), 759–771.

Hurvich, C. M., & Tsai, C.-L. (1989). Regression and time series model
selection in small samples. *Biometrika*, **76**(2), 297–307.

Schwarz, G. (1978). Estimating the dimension of a model.
*The Annals of Statistics*, **6**(2), 461–464.
