---
title: "Getting started with ICtoolkit"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Getting started with ICtoolkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

```{r load}
library(ICtoolkit)
data(covid_eol)
```

## Overview

ICtoolkit provides a unified interface for five information criteria across four
model classes:

| Criterion | Function | Description |
|-----------|----------|-------------|
| AIC | `compute_aic()` | Akaike (1974) |
| AICc | `compute_aicc()` | Small-sample correction |
| BIC | `compute_bic()` | Schwarz (1978) |
| EBIC | `compute_ebic()` | Extended BIC; Chen & Chen (2008) |
| RBIC | `compute_rbic()` | Grouped EBIC for structured predictors |

All functions support `lm`, `glm`, `glmnet`, and `ncvreg` objects.
For `glmnet` and `ncvreg`, the criterion is evaluated at every point on the
regularization path (one value per `lambda`), which can itself be used as a
selection algorithm without any stepwise fitting.

Throughout this vignette we use the `covid_eol` dataset bundled with the
package — a synthetic dataset modeled after a study of family members of
ICU patients surveyed for PTSD during the COVID-19 pandemic (n = 279).

---

## 1. Standard criteria for `lm` and `glm`

```{r lm-example}
fit <- lm(ies_total ~ age + race + hispanic + female, data = covid_eol)

compute_aic(fit)
compute_aicc(fit)
compute_bic(fit)
```

Each result is a scalar with informative attributes:

```{r attributes}
result <- compute_bic(fit)
attr(result, "fit_class")   # model class
attr(result, "k")           # number of estimated parameters
attr(result, "criterion")   # which IC
```

The same functions work for `glm` objects:

```{r glm-example}
fit_glm <- glm(ptsd ~ age + race + hispanic + female,
               data = covid_eol, family = binomial)
compute_aic(fit_glm)
compute_bic(fit_glm)
```

---

## 2. EBIC: Extended BIC

EBIC adds a sparsity-encouraging penalty to BIC:

$$\text{EBIC} = \text{BIC} + 2\,\gamma\,\log\binom{P}{k}$$

where $P$ is the total number of candidate predictors and $k$ is the number
selected.  The default $\gamma$ follows Chen & Chen (2008):

$$\gamma = \frac{\log(P/\sqrt{n})}{\log P} \cdot \mathbf{1}(P \geq \sqrt{n})$$

```{r ebic-lm}
# P = 10 candidate predictors (total, excluding intercept)
result_ebic <- compute_ebic(fit, P = 10)
result_ebic

attr(result_ebic, "gamma")   # the computed gamma
attr(result_ebic, "P")
```

You can supply your own `gamma`:

```{r ebic-custom-gamma}
# Fixed gamma = 0.5
compute_ebic(fit, P = 10, gamma = 0.5)
```

Setting `gamma = 0` recovers standard BIC:

```{r ebic-zero-gamma}
all.equal(
  as.numeric(compute_ebic(fit, P = 10, gamma = 0)),
  as.numeric(compute_bic(fit))
)
```

---

## 3. RBIC: Grouped EBIC for structured predictors

RBIC generalises EBIC to predictors with a natural group structure (e.g.,
main effects vs. interactions, or predictors from different data modalities):

$$\text{RBIC} = \text{BIC} + 2 \sum_{g=1}^{G} \gamma_g \log\binom{P_g}{k_g}$$

Groups are defined via a `P_index` list.  For `lm`/`glm` objects, each
element should be a character vector of **term labels** exactly as returned
by `attr(terms(fit), "term.labels")`.

```{r rbic-setup}
# Two groups: demographic variables and race/ethnicity variables
P_index <- list(
  demographic = c("age", "female"),
  race_eth    = c("race", "hispanic")
)

result_rbic <- compute_rbic(fit, P_index = P_index)
result_rbic

attr(result_rbic, "gamma")
```

---

## 4. IC over the regularization path with `glmnet`

For `glmnet` objects, all five criteria return a **vector** of length
`length(lambda)`.  This makes IC-based selection straightforward: pick the
`lambda` that minimises the criterion.

```{r glmnet-example, eval = requireNamespace("glmnet", quietly = TRUE)}
library(glmnet)

# Create a model matrix from the covid_eol predictors
X <- model.matrix(~ age + race + hispanic + female, data = covid_eol)[, -1]
y <- covid_eol$ies_total
p <- ncol(X)

fit_glmnet <- glmnet(X, y, family = "gaussian")

bic_path  <- compute_bic(fit_glmnet)
ebic_path <- compute_ebic(fit_glmnet, P = p)
rbic_path <- compute_rbic(fit_glmnet,
                           P_index = list(
                             demographic = c("age", "femaleYes"),
                             race_eth    = c("raceBlack", "raceAsian",
                                             "raceOther", "hispanicYes")))

# Lambda selected by each criterion
lambda_bic  <- attr(bic_path,  "lambda")[which.min(bic_path)]
lambda_ebic <- attr(ebic_path, "lambda")[which.min(ebic_path)]
lambda_rbic <- attr(rbic_path, "lambda")[which.min(rbic_path)]

cat("BIC-selected lambda: ",  round(lambda_bic,  4), "\n")
cat("EBIC-selected lambda:", round(lambda_ebic, 4), "\n")
cat("RBIC-selected lambda:", round(lambda_rbic, 4), "\n")

# Coefficients at BIC-optimal lambda
coef(fit_glmnet, s = lambda_bic)
```

> **Note on alpha < 1:** When a `glmnet` model is fit with `alpha < 1`
> (elastic-net or ridge), the package issues a warning because the count of
> nonzero coefficients may not be an appropriate measure of model complexity
> under L2 shrinkage.

---

## 5. Expanding scope with `expand_scope()`

When searching for interactions and polynomials, the number of candidate
terms grows quickly and constructing the `P_index` by hand becomes tedious.
`expand_scope()` automates this: given a base formula, it generates the
expanded formula and a `P_index` grouped by term type.

```{r expand-scope}
# Expand to pairwise interactions + quadratic terms
es <- expand_scope(ies_total ~ age + race + hispanic + female,
                   data = covid_eol, k = 2, poly = 2)
es$formula
es$P_index
```

---

## 6. Stepwise selection with `ic_step()`

`ic_step()` extends `MASS::stepAIC` to all five criteria for `lm`/`glm`
objects. Combined with `expand_scope()`, it makes RBIC-guided stepwise
selection through interactions straightforward.

```{r ic-step-example}
fit_null <- lm(ies_total ~ 1, data = covid_eol)

# Forward BIC selection through pairwise interactions + quadratic terms
selected <- ic_step(
  fit_null,
  scope     = list(lower = ~ 1, upper = es$formula),
  direction = "forward",
  criterion = "BIC",
  trace     = 1
)
formula(selected)
```

### Inspecting the selection path

`ic_step()` attaches a `step_path` data frame to the returned model:

```{r step-path}
attr(selected, "step_path")
```

---

## 7. Comparing criteria on a simulation

The ranked sparsity advantage of RBIC is most apparent when the candidate
set includes many derived variables (interactions, polynomials) that inflate
the search space. We illustrate with a simulation where the true model has
10 main-effect candidates but only 3 active main effects and 1 active
interaction — a sparse signal embedded in 65 candidate terms.

```{r sim-setup}
set.seed(42)
n <- 200; p <- 10
X <- matrix(rnorm(n * p), n, p,
            dimnames = list(NULL, paste0("x", 1:p)))
y <- 2*X[,1] - 1.5*X[,2] + X[,3] + 1.5*X[,1]*X[,2] + rnorm(n, sd = 2)
sim_data <- as.data.frame(cbind(y = y, X))

sim_es <- expand_scope(
  y ~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9 + x10,
  data = sim_data, k = 2, poly = 2
)
# 10 main + 10 quadratic + 45 pairwise = 65 candidates
sapply(sim_es$P_index, length)
```

We split into training and test sets, then compare backward stepwise
selection under AIC, BIC, and RBIC:

```{r comparison, warning = FALSE}
set.seed(2026)
train_idx <- sample(n, n %/% 2)
train <- sim_data[train_idx, ]
test  <- sim_data[-train_idx, ]

fit_full <- lm(sim_es$formula, data = train)
scope    <- list(lower = ~ 1, upper = sim_es$formula)

true_terms <- c("x1", "x2", "x3", "x1:x2")

criteria <- c("AIC", "BIC", "RBIC")
results  <- lapply(criteria, function(crit) {
  extra_args <- list()
  if (crit == "RBIC") extra_args <- list(P_index = sim_es$P_index)
  do.call(ic_step, c(
    list(object    = fit_full,
         scope     = scope,
         direction = "backward",
         criterion = crit,
         trace     = 0),
    extra_args
  ))
})
names(results) <- criteria

# Count terms by type, false positives, and test-set RMSPE
summary_tab <- t(sapply(results, function(m) {
  sel  <- attr(terms(m), "term.labels")
  pred <- predict(m, newdata = test)
  rmspe <- sqrt(mean((test$y - pred)^2))
  c(main       = sum(sel %in% sim_es$P_index$main),
    poly       = sum(sel %in% sim_es$P_index$poly_2),
    interact   = sum(sel %in% sim_es$P_index$interact_2),
    total      = length(sel),
    true_pos   = sum(sel %in% true_terms),
    false_pos  = sum(!sel %in% true_terms),
    RMSPE      = round(rmspe, 3))
}))
summary_tab
```

RBIC recovers the true model exactly — 4 true positives with 0 false
positives — and achieves by far the best prediction accuracy. BIC retains
many spurious interactions and polynomials despite its heavier penalty,
because it treats all 65 candidates with equal skepticism. RBIC's
group-specific penalties allow the true interaction (`x1:x2`) to enter
while correctly excluding the 44 noise interactions.

---

## References

Akaike, H. (1974). A new look at the statistical model identification.
*IEEE Transactions on Automatic Control*, **19**(6), 716–723.

Chen, J., & Chen, Z. (2008). Extended Bayesian information criteria for model
selection with large model spaces. *Biometrika*, **95**(3), 759–771.

Hurvich, C. M., & Tsai, C.-L. (1989). Regression and time series model
selection in small samples. *Biometrika*, **76**(2), 297–307.

Schwarz, G. (1978). Estimating the dimension of a model.
*The Annals of Statistics*, **6**(2), 461–464.
